---
title: "Examples for 'modexpress'"
date: March 23, 2026
output:
  github_document:
    fig_width: 10
    fig_height: 7
---
```{r include = FALSE}
library(modexpress)
```

## Examples of functions using test data (mostly iris or mtcars)**

### Starting with a simple linear model
```{r Simple univariate lm}
lm_1 <- stats::lm(Sepal.Length ~ Sepal.Width, data = iris)
```

You can express the fit by Species:
```{r lm_1_by_species, fig.alt="Expression of lm_1 by Species"}
express_fit(lm_1, iris, by_factor = "Species")
```

Or by by Petal.Length:
```{r lm_1_by_pl, fig.alt="Expression of lm_1 by Petal.Length"}
express_fit(lm_1, iris, by_covar = "Petal.Length")
```

Or you can express the fit by both at once:
```{r lm_1_by_both, fig.alt="Expression of lm_1 by Species and Petal.Length"}
express_fit(lm_1, iris, by_factor = "Species", by_covar = "Petal.Length")
```

Beyond that, we can plot model diagnostics by Species:
```{r lm_1_eval_by_species, fig.alt="Expression of lm_1 evaluation by Species"}
express_eval(lm_1, iris, by_factor = "Species")
```

Or by Petal.Length:
```{r lm_1_eval_by_pl, fig.alt="Expression of lm_1 evaluation by Petal.Length"}
express_eval(lm_1, iris, by_covar = "Petal.Length")
```

With `express_gauge()`, we can gauge potential missing parameters by comparing model residuals to another variable.
This will give you both a plot and a summary of a diagnostic test for that relationship.
For example, compare the model residuals for `lm_1` by Species:
```{r lm_1_gauge_by_species, fig.alt="Expression of lm_1 gauge by Species"}
express_gauge(lm_1, iris, by_factor = "Species")
```

Or gauge potential missing parameters by Petal.Length:
```{r lm_1_gauge_by_pl, fig.alt="Expression of lm_1 gauge by Petal.Length"}
express_gauge(lm_1, iris, by_covar = "Petal.Length")
```

Or, again, by both:
```{r lm_1_gauge_by_both, fig.alt="Expression of lm_1 gauge by Species and Petal.Length"}
express_gauge(lm_1, iris, by_factor = "Species", by_covar = "Petal.Length")
```

---

### More complex linear models
A univariate linear model is great, but what about multivariate linear models, those with interactions, and inline functions? Let's do a more complex model with `exp()`, `I(x**2)`, and interaction terms.
```{r More complex lm}
lm_2 <- stats::lm(
  Sepal.Length ~ exp(Sepal.Width) + I(Petal.Length ** 2) + Sepal.Width * Petal.Length,
  data = iris
)
```

We can now express partial fits for each term.
Expressing the partial fit by Species for one term, `exp(Sepal.Width)`:
```{r lm_2_by_species1, fig.alt="Expression of partial exp(Sepal.Width) fit by Species"}
express_one(lm_2, iris, "exp(Sepal.Width)", by_factor = "Species")
```

Or, we can express the partial fit by Petal.Width for another term, e.g., the interaction between Sepal.Width and Petal.Length:
```{r lm_2_by_pw1, fig.alt="Expression of partial SW:PL fit by Petal.Width"}
express_one(lm_2, iris, "Sepal.Width:Petal.Length", by_covar = "Petal.Width")
```

We can even do many terms at once, though this may crowd our plot if we have many terms:
```{r lm_2_by_species2, fig.alt="Expression of partial fits for many terms by Species"}
express_many(lm_2,
             iris,
             c("exp(Sepal.Width)", "Sepal.Width:Petal.Length"),
             by_factor = "Species")
```

Or, if we specify nothing, all terms at once:
The plot gets very busy with all those terms! If you want them individually, try `grid = FALSE`.
```{r lm_2_by_pw2, fig.alt="Expression of partial fits for all terms by Species"}
express_many(lm_2, iris, by_covar = "Petal.Width")
```

We can gauge potential missing parameters in the whole model by Species:
```{r lm_2_gauge_by_species, fig.alt="Expression of lm_2 gauge by Species"}
express_gauge(lm_2, iris, by_factor = "Species")
```

Or, now that we have multiple components, we can gauge potential missing parameters by partial residuals for one component, `exp(Sepal.Width)` by Species:
```{r lm_2_gauge_by_species2, fig.alt="Expression of lm_2 gauge for exp(SW) by Species"}
express_gaugepart(lm_2, iris, "exp(Sepal.Width)", by_factor = "Species")
```

Or, again, if we leave the components blank, we get everything (notice how long the summaries get once we use everything at once!). This time by Petal.Length:
```{r lm_2_gauge_by_pl, fig.alt="Expression of lm_2 gauge for all components by Species"}
express_gaugepart(lm_2, iris, by_covar = "Petal.Length")
```

---

### Simple univariate GAM
Note that the `control = list(keepData = TRUE)` in a GAM object allows you to not have to specify the original data every time. Let's start with a univariate GAM.
```{r Simple GAM}
gam_1 <- mgcv::gam(
  Sepal.Length ~ s(Sepal.Width),
  method = "REML",
  data = iris,
  control = list(keepData = TRUE)
)
```

Searching through GAM terms works by partial matching. Let's plot the component containing "Sepal" by Species:
```{r gam_1_by_species1, fig.alt="Expression of gam_1 Sepal by Species"}
express_one(gam_1, "Sepal", by_factor = "Species")
```

We can evaluate, too, by e.g., Petal.Width. Note that it will tell you that splitting histograms by a covariate is unsupported.
```{r gam_1_eval_by_pw, fig.alt="Expression of gam_1 evaluation by PW"}
express_eval(gam_1, by_covar = "Petal.Width")
```

### More complex GAM
How about a GAM with more terms? Note the control parameter as above.
```{r Complex GAM}
gam_2 <- mgcv::gam(
  Sepal.Length ~ s(Sepal.Width) + s(Petal.Length) + ti(Sepal.Width, Petal.Length),
  method = "REML",
  data = iris,
  control = list(keepData = TRUE)
)
```

It is possible to express single and multivariate smooths. For example, express the s(Petal.Length) term by Species:
```{r gam_2_by_species1, fig.alt="Expression of gam_2 s(Petal.Length) by Species"}
express_one(gam_2, "s,Petal", by_factor = "Species")
```

The ti(Sepal.Width, Petal.Length) term is bivariate. This can also be expressed:
```{r gam_2_by_species2, fig.alt="Expression of gam_2 ti(Sepal.Width, Petal.Length) by Species"}
express_one(gam_2, "ti,Se,Pe", by_factor = "Species")
```

Or, again, if left blank, it will express everything (here, by Petal.Width):
```{r gam_2_by_pw, fig.alt="Expression of all gam_2 components by PW"}
express_many(gam_2, by_covar = "Petal.Width")
```

We can gauge potential missing parameters by Species:
```{r gam_2_gauge_by_species, fig.alt="Expression of gam_2 gauge by Species"}
express_gauge(gam_2, by_factor = "Species")
```

Or, by component, we can gauge potential missing parameters by partial residuals:
```{r gam_2_gauge_by_species2, fig.alt="Expression of gam_2 gauge s(Sepal.Width) by Species"}
express_gaugepart(gam_2, "s,Se", by_factor = "Species")
```

And again for many or all components:
```{r gam_2_gauge_by_species3, fig.alt="Expression of gam_2 gauge all components by Species"}
express_gaugepart(gam_2, by_factor = "Species")
```

### GAM with random effect smooth (hierarchical GAM)
```{r Hierarchical GAM}
gam_3 <- mgcv::gam(
  Sepal.Length ~ s(Sepal.Width) + s(Species, bs = "re"),
  method = "REML",
  data = iris,
  control = list(keepData = TRUE)
)
```

Let's see how the random effect gets plotted. What happens here is that the mean value for a continuous variable is taken for each level in the random effect.
```{r gam_3_by_pw_1, fig.alt="Expression of gam_3 s(Species) by Petal.Width"}
express_one(gam_3, "Spec", by_covar = "Petal.Width")
```

We can do model evaluations on GAMs:
```{r gam_3_eval_by_pw, fig.alt="Expression of gam_3 evaluation by Petal.Width"}
express_eval(gam_3, by_covar = "Petal.Width")
```

And gauge potential missing effects:
```{r gam_3_gauge_by_pw, fig.alt="Expression of gam_3 gauge by Petal.Width"}
express_gauge(gam_3, by_covar = "Petal.Width")
```

### GAM with a non-tensor multivariate smooth
For a GAM with a non-tensor multivariate smooth, note that it essentially works the same as the tensor product smooths.
```{r Non-tensor product GAM}
gam_4 <- mgcv::gam(
  Sepal.Length ~ s(Sepal.Width, Petal.Length),
  method = "REML",
  data = iris,
  control = list(keepData = TRUE)
)
```

This is what that looks like:
```{r gam_4_express, fig.alt="Expression of gam_4 components"}
express_many(gam_4)
```

We can also do this for factor smooths (`bs = "fs"`):
```{r Factor smooth GAM}
gam_5 <- mgcv::gam(
  Sepal.Length ~ s(Sepal.Width, Species, bs = "fs"),
  method = "REML",
  data = iris,
  control = list(keepData = TRUE)
)
```

Note how the factor smooth will show a separate effect for each level of the smooth:
```{r gam_5_express, fig.alt="Expression of gam_5 components"}
express_many(gam_5)
```

The model diagnostics look mostly the same:
```{r gam_5_eval, fig.alt="Expression of gam_5 evaluation"}
express_eval(gam_5)
```

---

### Simple GLMs
Just to test, we can create an inverse-gaussian log-linked GLM. This is probably not the right fit for the data, but it is useful for testing.
```{r Inverse gaussian GLM}
glm_1 <- stats::glm(Sepal.Length ~ Sepal.Width,
                    data = iris,
                    family = inverse.gaussian(link = "log"))
```

We can also create one with a binomial logit-linked GLM, because that will probably look quite different.
```{r Binomial GLM}
glm_2 <- stats::glm(vs ~ wt + disp,
                    data = mtcars,
                    family = binomial(link = "logit"))
```

GLM objects contain the full data on which they were built. This means we do not need to pass the `data` parameter at all. This means however that we do need to feed something to `data` when making the model.

Note that the shapes are off for a lot of the plots for a binomial distribution. However, this is also true for the base `plot.glm` method.
```{r glm_2_eval, fig.alt="Expression of glm_2 evaluation"}
express_eval(glm_2)
```

```{r glm_1_gauge_by_species, fig.alt="Expression of glm_1 gauge by Species"}
express_gauge(glm_1, by_factor = "Species")
```

### More complex GLMs
Let's try the same inverse-gaussian GLM, but with more terms, like an interaction between Sepal.Width and Petal.Length.
```{r Complex GLM}
glm_3 <- stats::glm(Sepal.Length ~ Sepal.Width * Petal.Length,
                    data = iris,
                    family = inverse.gaussian(link = "log"))
```

Expressing all components by Species:
```{r glm_3_by_species, fig.alt="Expression of glm_3 components by Species"}
express_many(glm_3, by_factor = "Species")
```

Gauging by individual parts for the complex model, by Species and Petal.Width:
It will tell us that the fit is singular, and that comes from an LMM that is actually being fit to show the correlation between Petal.Width and our model residuals, with a random slope and intercept of Species.
```{r glm_3_gauge_by_species, fig.alt="Expression of glm_3 gauge by Species"}
express_gaugepart(glm_3, by_factor = "Species", by_covar = "Petal.Width")
```

### Linear mixed models
We could try a linear mixed model with a random intercept:
```{r Random intercept LMM}
lmm_1 <- lme4::lmer(Sepal.Length ~ Sepal.Width + (1|Species),
                    data = iris)
```

Or, with a random intercept and a random slope:
```{r Random intercept/slope LMM}
lmm_2 <- lme4::lmer(Sepal.Length ~ Sepal.Width + (Sepal.Width|Species),
                    data = iris)
```

Expressing a linear term:
```{r lmm_1_sw, fig.alt="Expression of lmm_1 Sepal.Width"}
express_one(lmm_1, iris, "Sepal.Width")
```

Expressing a random term (note that it is not "(1|Species)", just "Species"):
```{r lmm_1_species, fig.alt="Expression of lmm_1 Species"}
express_one(lmm_1, iris, "Species")
```

Expressing a random term with both an intercept and slope:
```{r lmm_2_species, fig.alt="Expression of lmm_2 Species"}
express_one(lmm_2, iris, "Species")
```

Expressing all components:
```{r lmm_2_all, fig.alt="Expression of lmm_2 components"}
express_many(lmm_2, iris)
```

For LMMs, `express_eval()` is similar to the GAM method note that `plot.lmerMod` returns a single plot of fitted values vs. residuals, that is equivalent to `express_linpred()` and is included in the evaluation.
```{r lmm_2_eval_by_pl, fig.alt="Expression of lmm_2 evaluation by PW"}
express_eval(lmm_2, iris, by_covar = "Petal.Length")
```

Gauging a full model:
```{r lmm_2_gauge_by_pl, fig.alt="Expression of lmm_2 gauge by PW"}
express_gauge(lmm_2, iris, by_covar = "Petal.Length")
```

Gauging with LMMs is tricky because only the linear terms have partial residuals. Instead, for random effect terms, the effects are compared between another covariate or a factor, if the factor has fewer levels than there are random effects.
```{r lmm_2_gauge_species_by_pl, fig.alt="Expression of lmm_2 gauge Species by PL"}
express_gaugepart(lmm_2, iris, "Species", by_covar = "Petal.Length")
```

---

### Generalized linear mixed models
```{r A simple GLMM}
glmm_1 <- lme4::glmer(Sepal.Length ~ Sepal.Width + (1|Species),
                      data = iris,
                      family = inverse.gaussian(link = "log"))
```

Expressing one GLMM component:
```{r glmm_sw_by_pl, fig.alt="Expression of glmm_1 Sepal.Width by PL"}
express_one(glmm_1, iris, "Sepal.Width", by_covar = "Petal.Length")
```

Expressing many GLMM components:
```{r glmm_all_by_pl, fig.alt="Expression of glmm_1 components by PL"}
express_many(glmm_1, iris, by_covar = "Petal.Length")
```

Evaluating a GLMM object by Species:
```{r glmm_eval_by_species, fig.alt="Expression of glmm_1 evaluation by Species"}
express_eval(glmm_1, iris, by_factor = "Species")
```

Gauging a full model:
```{r glmm_1_gauge, fig.alt="Expression of glmm_1 gauge by Species"}
express_gauge(glmm_1, iris, by_factor = "Species")
```

Gauging a single model component:
```{r glmm_1_gauge_sw_by_pw, fig.alt="Expression of glmm_1 gauge Sepal.Width by PW"}
express_gaugepart(glmm_1, iris, "Sepal.Width", by_covar = "Petal.Width")
```
